---
layout: 12pt
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \usepackage{todonotes}
  - \doublespacing
  - \usepackage{rotating}
  - \usepackage{color, soul}
  - \usepackage[font={normalsize},labelfont={bf},labelsep=quad]{caption}
  - \usepackage{sectsty}
  - \usepackage{bm,mathrsfs}
  - \usepackage{mathptmx} 
bibliography: ~/Dropbox/Bibliography/usses_water.bib
csl: components/ecology_letters.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in
linkcolor: black
urlcolor: black
---

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
\definecolor{blue}{rgb}{0,0,0.7} 
\newcommand{\new}{\textcolor{blue}}

\begin{singlespace}

\begin{centering}

\Large{\textbf{Ecosystem and community resistance to six years of drought and deluge in the Sagebrush Steppe}}

\bigskip{} \bigskip{}

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

\normalsize{Andrew T. Tredennick\textsuperscript{1}, Peter B. Adler\textsuperscript{1}, and Andrew R. Kleinhesselink\textsuperscript{2}}

\bigskip{}

\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}}
\textit{\small{\textsuperscript{2}Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California xxxxx}}

\end{centering}

\vspace{3em}

Last compile: \today

\end{singlespace}

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "manuscript"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```


```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

\begin{abstract}
Summarize it here
\vspace{2em}
\end{abstract}


\setlength{\parindent}{5ex}

Introduction
------------
Start it here

Methods
-------
### Study Area
We conducted our precipitation manipulation experiment at the United States Sheep Experimental Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The vegetation is typical of high elevation sagebrush steppe.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment, average mean annual precipitation was xxx mm and mean monthly temperature ranged from -x$^{\circ}$C in January to x$^{\circ}$C in July.

### Precipitation Experiment
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we (well, one of us [P. Adler]) relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots that have recieved no treatment, just ambient precipitation.
In spring 2011, we (well, two of us [A. Kleinhesselink and P. Adler]) established 16 new 1 m$^2$ plots.
We avoided areas on steep hill slopes, areas with greater than 20\% cover of bare rock, and areas with greater than 10\% cover of the shrubs *Purshia tridentata* and/or *Amelanchier utahensis*.
We established the new plots in pairs and randomly assigned each plot in a pair to receive a "drought" or "irrigation" treatment.

Drought and irrigation treatments were designed to decrease and increase the amount of ambient precipitation by 50\%, respectively.
To achieve this, we used a system of rain-out shelters and automatic irrigation [@Gherardi2013].
The rain-out shelters consisted of transparent acrylic shingles 1-1.5 m above the ground that covered an area of 2.5 by 2 m.
The shingles intercepted approximately 50\% of incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on to the adjacent irrigation plot via two suspended sprinklers.
Pumping was triggered by float switches once water levels reached about 20 liters.
We disconnected the irrigation pumps each fall and reconnected them, often with difficulty, each spring.
The rain-out shelters remained in place throughout the year.

To make sure the treatments were having the desired effects, we monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pulman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors in each plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06/06/2012, 04/29/2015, 05/07/2015, 06/09/2015, and 05/10/2016 using a handheld EC-5 sensor.

### Data Collection
#### Aboveground Net Primary Productivity
We estimated aboveground net primary productivity (ANPP) using a radiometer to related ground reflectance to plant biomass [see @Byrne2011 for a review].

Acknowledgments
---------------
This work was funded by the National Science Foundation through grants DEB-1353078 and DEB-1054040 to PBA, a Graduate Research Fellowship to ARK, and a Postdoctoral Resaerch Fellowship in Biology to ATT (DBI-1400370). 
Additional support came from the Utah Agricultural Experiment Station (journal paper xxxx).
We thank the many summer research technicians who collected the data reported in this paper.

\singlespace
\setlength{\parindent}{0ex}

References
----------
