---
layout: 12pt
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \usepackage{todonotes}
  - \doublespacing
  - \usepackage{rotating}
  - \usepackage{color, soul}
  - \usepackage{sectsty}
  - \usepackage{bm,mathrsfs}
  - \usepackage{mathptmx} 
bibliography: ~/Dropbox/Bibliography/usses_water.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in
linkcolor: black
urlcolor: black
---

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
\definecolor{blue}{rgb}{0,0,0.7} 
\newcommand{\new}{\textcolor{blue}}

\begin{singlespace}

\begin{centering}

\Large{\textbf{Ecosystem and community resistance to five years of drought and deluge in a sagebrush tteppe}}

\bigskip{} \bigskip{}

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

\normalsize{Andrew T. Tredennick\textsuperscript{1}, Andrew R. Kleinhesselink\textsuperscript{2}, and Peter B. Adler\textsuperscript{1}}

\bigskip{}

\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}}
\textit{\small{\textsuperscript{2}Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California xxxxx}}

\end{centering}

\vspace{3em}

Last compile: \today

\end{singlespace}

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "manuscript"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```


```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

\begin{abstract}
Precipitation is predicted to become more variable in the western U.S., meaning years of above and below average precipitation will become more common.
Such periods of drought and deluge could become major drivers of plant community dynamics and ecosystem functioning.
We used drought and irrigation treatments (50\% decrease/increase) to see how a sagebrush steppe plant community in Idaho will respond to future climate changes.
The ecosystem was suprisingly resistant to both drought and irrigation.
Aboveground net primary productivity (ANPP) and its temporal stability did no differ among treatments.
The similarity of ecosystem functioning was not due to shifts at the plant community level, where species compositions and species synchrony, an important driver of ecosystem stability, among treatments were similar.
However, ANPP does appear to be diverging among treatments with time, even though the effect is not significant.
At least in the short-term, this sagebrush steppe ecosystem is resistant to increases and decreases in annual precipitation.

\vspace{2em}

\emph{Key words: drought, aboveground net primary productivity, ecosystem stability, climate change, species synchrony, sagebrush steppe}
\end{abstract}


\setlength{\parindent}{5ex}

Introduction
------------
Chronic alterations to resource availability should cause community composition to shift over time.
Species' relative abundances are expected to re-order and, eventually, some species will be lost from the local community and new species will be gained [@Smith2009; @Avolio2015].
Ecosystem functioning, such as productivity and stability, may remain relatively constant (CITATION) or shift to a new mean state (CITATION) depending on the trait-space of the new community.

Different ecosystems are sensitive to different global changes.
For example, @Flombaum2017 found aboveground net primary productivity (ANPP) in a Patagonian steppe was less sensitive to drought than a North American tallgrass prairie, relative to other global changes.
Taken together with other results, @Flombaum2017 conclude that ecosystem sensitivity to a resource alteration is inversly related to the natural variation of the resource.
In the above example, precipitation is much more variable in the Patagonian steppe than the North American tallgrass prairie.

This leaves us with the following qualitative understanding.
Chronic resource alteration should induce shifts in community composition and, maybe, ecosystem functioning, but only if the focal ecosystem is predisposed to be sensitive to the particular resource that is altered.
While a crude understanding, this does allow us to make testable predictions when confronted with an ecosystem--resource pairing.

In many areas of the western United States, precipitation is likely to become more variable.
This will result in swings back-and-forth between multiyear periods of above and above average precipitation.
An obvious question is, how will semiarid ecosystems respond to periods of drought and deluge?
A naive expectation is that alterations in precipitation, and subsequently available soil moisture, should cause changes in plant community composition and increase or decrease ANPP dependending on the direction of precipitation change.
But our emerging qualitative understanding suggests an alternative expectation: altering soil moisture may have little to no effect on plant community composition and ecosystem functioning.
This expectation stems from the fact that precipitation is already a variable resource in semiarid systems, meaning the plant community is not sensitive to realistic increases/decreases in precipitation.

Here we test our qualitative understanding of how altering precipitation will impact a sagebrush steppe ecosystem by imposing drought and irrigation for five years.
In particular, we test the following competing predictions:

\begin{description}
\item [P1.] Altering precipitation will favor certain species over others, resulting in a shift in the plant community composition. ANPP will increase (irrigation) or decrease (drought), and the treatment effects will get stronger over time (a treatment$\times$year interaction).

\item[P2.] Altering precipitation will not favor certain species and plant community composition will not change. ANPP will increase (irrigation) or decrease (drought), and the treatment effects will get stronger over time.
\end{description}

\noindent{}Note that both predictions assume ANPP will respond to precipitation alteration.
Under **P1**, ANPP response stems directly from plant community shifts to species able to take advantage of the new precipitation regime.
Under **P2**, ANPP response stems from all species responding similarly to changes in precipitation.



Methods
-------
### Study Area
We conducted our precipitation manipulation experiment at the United States Sheep Experimental Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The vegetation is typical of high elevation sagebrush steppe.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment, average mean annual precipitation was xxx mm and mean monthly temperature ranged from -x$^{\circ}$C in January to x$^{\circ}$C in July.

### Precipitation Experiment
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we (well, one of us [P. Adler]) relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots that have recieved no treatment, just ambient precipitation.
In spring 2011, we (well, two of us [A. Kleinhesselink and P. Adler]) established 16 new 1 m$^2$ plots.
We avoided areas on steep hill slopes, areas with greater than 20\% cover of bare rock, and areas with greater than 10\% cover of the shrubs *Purshia tridentata* and/or *Amelanchier utahensis*.
We established the new plots in pairs and randomly assigned each plot in a pair to receive a "drought" or "irrigation" treatment.

Drought and irrigation treatments were designed to decrease and increase the amount of ambient precipitation by 50\%, respectively.
To achieve this, we used a system of rain-out shelters and automatic irrigation [@Gherardi2013].
The rain-out shelters consisted of transparent acrylic shingles 1-1.5 m above the ground that covered an area of $2.5\times2$ m.
The shingles intercepted approximately 50\% of incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on to the adjacent irrigation plot via two suspended sprinklers.
Pumping was triggered by float switches once water levels reached about 20 liters.
We disconnected the irrigation pumps each fall and reconnected them, often with difficulty, each spring.
The rain-out shelters remained in place throughout the year.

To make sure the treatments were having the desired effects, we monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pulman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors in each plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06/06/2012, 04/29/2015, 05/07/2015, 06/09/2015, and 05/10/2016 using a handheld EC-5 sensor.

### Data Collection
#### Aboveground Net Primary Productivity
We estimated aboveground net primary productivity (ANPP) using a radiometer to relate ground reflectance to plant biomass [see @Byrne2011 for a review].
We recorded ground reflectance at four wavelengths, two associated with red reflectance (xxx and xxx) and two associated with near-infrared reflectance (xxx and xxx).
At each plot in each year, we took four readings of ground reflectances at the above wavelengths.
We also took readings in ten calibration plots adjacent to the experimental site, in which we harvested all aboveground biomass, dried it to a constant weight at 60$^{\circ}$C, and weighed it to estimate ANPP.

For each plot and year, we averaged the four readings for each wavelength and then calculated NDVI using the MODIS and AVHRR algorithms.
To convert NDVI to ANPP we regressed NDVI against the dry biomass weight from the ten calibration plots.
We fit regressions to MODIS-based NDVI and AVHRR-based NDVI for each year and retained the regression with the better fit.
Using the best regression equation for each year, we predicted ANPP.

#### Species Composition
Species composition data came from annual census maps for each plot made using a pantograph [@Hill1920].
The maps record the spatial location and size of each individual plant.
Using those annual maps, we aggregated over individuals to calculate total basal cover for each species in each plot.

### Data Analysis
####  ANPP and Ecosystem Stability
We used ANOVA with treatment, year, and treatment$\times$year terms to test the effects of the treatments on ANPP over the course of the five year experiment.
To test for within-year differences of ANPP among treatments, we fit ANOVAs independently for each year with treatment terms only.
We used the \texttt{lm()} function in \texttt{R} [v3.3.2, @R2016] to fit the models, and the \texttt{Anova()} function from the \texttt{car} package [@Fox2011] to extract sums-of-squares tables with Type II tests.

We calculated ecosystem stability as the temporal mean of ANPP divided by its temporal standarad deviation $\left( \overline{\text{ANPP}} / \sigma(\text{ANPP}) \right)$.
Ecosystem stability was calculated for each plot.
We compared ecosystem stability among treatments using ANOVA (same \texttt{R} functions as above).

####  Species Synchrony and Species Composition
We calculated the synchrony of species fluctuations through time using basal area cover data and individual density data.
For both data types we calculated the synchrony metric described by @Loreau2008a, which ranges from 0 (perfect ascynchrony) to 1 (perfect synchrony).
Before calculating synchrony, we averaged cover and density over plots within treatments for each species.
We used the \texttt{synchrony()} function in the \texttt{R} package \texttt{codyn} [@Hallett2016] to calculate species synchrony.

To see if community composition differed among treatments through time, we used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis distances.
For each year of the experiment, we first calculated Bray-Curtis distances among all plots, and then extracted those distances for use in the NMDS.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used functions in the \texttt{R} package \texttt{vegan} [@Oksanen2016] to calculate Bray-Curtis distances and the run the NMDS analysis.

Results
-------

Discussion
----------


Acknowledgments
---------------
This work was funded by the National Science Foundation through grants DEB-1353078 and DEB-1054040 to PBA, a Graduate Research Fellowship to ARK, and a Postdoctoral Resaerch Fellowship in Biology to ATT (DBI-1400370). 
Additional support came from the Utah Agricultural Experiment Station (journal paper xxxx).
We thank the many summer research technicians who collected the data reported in this paper.

\singlespace
\setlength{\parindent}{0ex}

\newpage{}

Tables
------

```{r table, include=TRUE, comment=FALSE,warning=FALSE,message=FALSE,results='asis'}
library(xtable)
# Read in within-year ANOVA results table
inyear_anova <- readRDS("../results/within_year_anova_table.RDS")

inyear_caption <- "ANOVA for treatment effects on ANPP for each year of the experiment."
print(xtable(inyear_anova, caption = inyear_caption, digits = 2),
      caption.placement="top",
      include.rownames = F, 
      sanitize.colnames.function = identity,
      comment = FALSE,
      size = "normalsize")
```

\newpage{}

Figures
-------

\begin{figure}[!ht]
  \centering
      \includegraphics[width=6in]{../figures/histppt_and_anpptrend.png}
  \caption{(A) Histogram of historical precipitation from 1926-2016, with the years of the experiment overlaid as colored points. ``Drought'' treatments (squares) are shown as a 50\% reduction from the observed precipitation. ``Irrigation'' treatment precipitation levels are not shown. (B) Mean (points) ANPP and its standard deviation (error bars) for each year of the experiment.}
\end{figure}


\newpage{}

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/nmds_comp_combined.png}
  \caption{Nonmetric multidimensional scaling scores representing plant communities in each plot, colored by treatment. (A) NMDS results using basal cover data; (B) NMDS results using individual density data. 2011 is a pre-treatment year.}
\end{figure}


\newpage{}

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/rank_clocks_cover.png}
  \caption{Rank clocks for cover by treatment. (A) All species. (B) Dominant species.}
\end{figure}


\newpage{}

References
----------
